---
title: "RcppTrust: A Thread-Safe C++ Trust Region Optimizer"
author: "Matthew Fidler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RcppTrust: A Thread-Safe C++ Trust Region Optimizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`RcppTrust` is a C++ port of the trust-region optimizer in Charles J.
Geyer's CRAN package `trust`, written for use inside `nlmixr2est`. This
vignette does not re-derive the algorithm -- see Geyer's original
paper/vignette, or Nocedal and Wright (1999, Chapter 4) and Fletcher
(1987, Section 5.1), for that. Instead it covers what a user of either
package actually needs to know: what carries over unchanged, what's
new, and how to call the new thread-safe C interface from your own
package.

## Similarities: what carries over unchanged

`RcppTrust::trust()` is a drop-in replacement for `trust::trust()`. It
has the same arguments, the same defaults, the same return value, and
implements the *exact same algorithm*: the same Newton /
easy-easy / hard-easy / hard-hard case split for the trust-region
subproblem, the same accept/reject and radius-adjustment rule, and the
same termination criteria.

```{r}
library(RcppTrust)

# Rosenbrock's function, exactly the example from ?trust::trust
objfun <- function(x) {
  f <- expression(100 * (x2 - x1^2)^2 + (1 - x1)^2)
  g1 <- D(f, "x1"); g2 <- D(f, "x2")
  h11 <- D(g1, "x1"); h12 <- D(g1, "x2"); h22 <- D(g2, "x2")
  x1 <- x[1]; x2 <- x[2]
  list(
    value = eval(f), gradient = c(eval(g1), eval(g2)),
    hessian = rbind(c(eval(h11), eval(h12)), c(eval(h12), eval(h22)))
  )
}

out <- trust(objfun, c(3, 1), 1, 5)
out[c("value", "argument", "converged", "iterations")]
```

Everything you already know about `trust::trust()` -- the `objfun`
contract (return a list with `value`, `gradient`, `hessian`, or
`list(value = Inf)`/`list(value = -Inf)` to signal an infeasible
point), the `rinit`/`rmax`/`parscale`/`iterlim`/`fterm`/`mterm`/
`minimize`/`blather` arguments, the shape of the returned list,
`blather = TRUE`'s extra `argpath`/`steptype`/`rho`/... components --
applies unchanged. All 10 of upstream's own `tests/*.R` scripts are
ported into this package's test suite and pass against the C++ core.

## Differences

| | `trust` | `RcppTrust` |
|---|---|---|
| Implementation | pure R | C++ (RcppArmadillo for the linear algebra) |
| Root-finder for the trust-region subproblem | `stats::uniroot()` | `boost::math::tools::toms748_solve()` |
| Callable from R | yes | yes, identically (`trust()`) |
| Callable from thread-safe C/C++ | no | yes (`trust_solve_c()`, see below) |
| Usable from another package without linking against its shared library | no C API at all | yes, via a header-only function-pointer table |
| Authors | Charles J. Geyer | Charles J. Geyer (algorithm), Matthew Fidler (C++ port) |

A few of these are worth expanding on:

- **Numeric agreement, not bit-identical output.** Swapping
  `uniroot()` for `toms748_solve()` (both are derivative-free,
  bracketing root-finders, just different implementations) means
  results agree with upstream to root-finder tolerance
  (`.Machine$double.eps^0.25`, upstream's own default) rather than to
  the last bit -- typically 1e-4 to 1e-8 in the returned `argument`,
  accumulating slightly over iterations. Convergence, iteration count,
  and the sequence of `steptype`s match exactly.

- **The hard-hard case can mirror-flip a sign.** When the trust-region
  subproblem hits an exact eigenvalue tie (the "hard-hard" case), the
  solution involves an eigenvector that's only defined up to sign.
  RcppTrust uses RcppArmadillo/LAPACK's `eig_sym()` rather than R's
  `eigen()`, which can pick the opposite sign than upstream did. This
  produces a mirror-image (equally valid) solution with the *same*
  objective value, convergence, and iteration count -- just a flipped
  sign on the affected component. This is a property of
  eigendecomposition at exact ties, not a bug in the port.

- **Thread safety is new.** `trust::trust()` calls back into R on
  every iteration and so can only ever run on R's main thread. Nothing
  about `RcppTrust::trust()` changes that -- it *also* calls back into
  R for the objective function, and is not more thread-safe than
  upstream. What's new is a second, separate entry point,
  `trust_solve_c()`, that never touches R at all and is safe to call
  from parallel C++ code (verified with an OpenMP stress test in this
  package's own test suite).

## When to use which interface

- **Writing an R script, or need a drop-in replacement for
  `trust::trust()`?** Use `RcppTrust::trust()`. It's a straight
  substitute.
- **Calling from your own C/C++ code, especially from inside a
  parallel loop (OpenMP, or similar)?** Use the C interface,
  `trust_solve_c()`, described below. This is the intended use in
  `nlmixr2est`, e.g. fitting each subject's inner problem on its own
  thread.
- **Writing a package that wants to call `trust_solve_c()` without
  taking on a hard link-time dependency on RcppTrust's shared
  library?** Use the header-only registration mechanism, also
  described below -- it's how the C interface is actually meant to be
  consumed by another package.

## Using the C interface

The thread-safe entry point is declared in
`inst/include/trust_types.h` (a plain C header with no R and no C++
dependency, so it's safe to include from plain C code too):

```c
typedef int (*trust_c_objfun_t)(int n, const double *par,
    double *value, double *gradient, double *hessian, void *userdata);
```

Your objective function fills `*value` (and `gradient`/`hessian`, each
length `n`/`n*n`, unless the point is infeasible) and returns:

- `0` -- feasible: `*value`, `gradient`, `hessian` are all filled and
  finite.
- `1` -- infeasible/out of the objective's domain: only `*value` is
  set, to `+INFINITY` when minimizing or `-INFINITY` when maximizing
  (`gradient`/`hessian` are left untouched). This is the C-level
  equivalent of upstream's `list(value = Inf)` convention.
- negative -- a hard error (the evaluation could not be completed at
  all).

Options are a plain struct, built from `trust_options_default()` (which
fills in `trust()`'s R-level defaults) and then adjusted as needed:

```c
trust_options_t opts = trust_options_default(/* rinit = */ 1.0, /* rmax = */ 5.0);
opts.iterlim = 200;      // default 100
opts.minimize = 0;       // maximize instead
opts.blather = 1;        // collect per-iteration arrays, like blather = TRUE
opts.has_parscale = 1;
opts.parscale = my_parscale;  // a `const double *` of length n you own
```

The result is written into a caller-supplied `trust_result_t`, which
owns its buffers (`argument`, `gradient`, `hessian`, and -- if
`opts.blather` -- the `argpath`/`argtry`/`steptype`/`rho`/... arrays,
laid out exactly like the R-level `blather = TRUE` output) until you
free it:

```c
trust_result_t res;
int rc = trust_solve_c(n, parinit, my_objfun, my_userdata, &opts, &res);
// rc == 0 on success; res.converged, res.iterations, res.argument, ...
trust_result_free(&res);
```

`trust_solve_c()` and `trust_result_free()` take no locks and touch no
shared, mutable state -- every argument is either an input or owned
exclusively by the caller's `trust_result_t` -- so many threads can
each be running their own `trust_solve_c()` call at the same time, on
independent problems, with no coordination needed. That's the piece
`trust::trust()` cannot offer at all.

### Calling it from your own package: the registration pattern

`RcppTrust`'s shared library is never linked against directly. Instead
-- following the same header-only, positionally-indexed
function-pointer pattern already used across the nlmixr2 ecosystem
(`rxode2`, `n1qn1`, `lbfgsb3c`) -- a consumer resolves
`trust_solve_c`/`trust_result_free` as function pointers once, at load
time. The full wiring is three pieces, which the example below
reproduces in a single self-contained file for demonstration.

1. Add `RcppTrust` to your `DESCRIPTION`'s `LinkingTo` (for the header)
   and `Imports` (for the R-level `.RcppTrustPtr()` getter).
2. In one translation unit, include the registration header with a
   package-unique renaming `#define` so the generated init function
   doesn't collide with any other consumer's:

   ```cpp
   extern "C" {
   #define iniRcppTrustPtrs _mypkg_iniRcppTrustPtrs
   #include <RcppTrust.h>
   iniRcppTrust
   }
   ```

3. In your package's `.onLoad()`, resolve the pointers once:

   ```r
   .onLoad <- function(libname, pkgname) {
     .Call(`_mypkg_iniRcppTrustPtrs`, RcppTrust:::.RcppTrustPtr(), PACKAGE = "mypkg")
   }
   ```

   After that, `trust_solve_c_ptr` and `trust_result_free_ptr` are
   live function pointers with the same signatures as
   `trust_solve_c()`/`trust_result_free()`, usable anywhere in your
   package's C++ -- including inside an OpenMP loop fitting many
   subjects' problems in parallel:

   ```cpp
   #pragma omp parallel for
   for (int i = 0; i < nSubjects; i++) {
     trust_options_t opts = trust_options_default(1.0, 5.0);
     trust_result_t res;
     trust_solve_c_ptr(n, parinit[i], subject_objfun, &subjectData[i], &opts, &res);
     // ... use res.argument, res.value, res.converged ...
     trust_result_free_ptr(&res);
   }
   ```

The example below is the whole pattern collapsed into one file and run
live, via `Rcpp::sourceCpp()`, exactly as it would work split across a
real package's `src/init.c`, `R/zzz.R`, and wherever the fit happens:

```{r}
cpp_code <- '
// [[Rcpp::depends(RcppTrust)]]
#include <Rcpp.h>

extern "C" {
#define iniRcppTrustPtrs _vignette_iniRcppTrustPtrs
#include <RcppTrust.h>
iniRcppTrust
}

// A thread-safe C objective function: Rosenbrock again, this time
// filling value/gradient/hessian directly instead of returning a list.
extern "C" int rosenbrock_c(int n, const double *par, double *value,
                             double *gradient, double *hessian, void *ud) {
  double x1 = par[0], x2 = par[1];
  double t = x2 - x1 * x1;
  *value = 100.0 * t * t + (1.0 - x1) * (1.0 - x1);
  gradient[0] = -400.0 * x1 * t - 2.0 * (1.0 - x1);
  gradient[1] = 200.0 * t;
  hessian[0] = -400.0 * x2 + 1200.0 * x1 * x1 + 2.0;
  hessian[1] = hessian[2] = -400.0 * x1;
  hessian[3] = 200.0;
  return 0;
}

// [[Rcpp::export]]
Rcpp::List fit_rosenbrock(SEXP ptrTable, Rcpp::NumericVector parinit) {
  // normally done once, in .onLoad() -- see above
  _vignette_iniRcppTrustPtrs(ptrTable);

  trust_options_t opts = trust_options_default(1.0, 5.0);
  trust_result_t res;
  trust_solve_c_ptr(parinit.size(), parinit.begin(), rosenbrock_c, nullptr, &opts, &res);

  Rcpp::List out = Rcpp::List::create(
    Rcpp::_["argument"] = Rcpp::NumericVector(res.argument, res.argument + res.n),
    Rcpp::_["value"] = res.value,
    Rcpp::_["converged"] = res.converged != 0,
    Rcpp::_["iterations"] = res.iterations);
  trust_result_free_ptr(&res);
  return out;
}
'
Rcpp::sourceCpp(code = cpp_code)

fit_rosenbrock(RcppTrust:::.RcppTrustPtr(), c(3, 1))
```

Note that this compiled file never links against `RcppTrust`'s shared
library at all -- `LinkingTo`/`Rcpp::depends()` only adds its
`inst/include` directory to the compiler's include path. Every
`trust_*` symbol used above except the two resolved pointers
(`trust_solve_c_ptr`, `trust_result_free_ptr`) is a type or macro, not
a function call, which is exactly the point: nothing here creates an
ABI dependency on a specific build of RcppTrust, only on the stable,
append-only pointer table it publishes.
