---
title: "More Trust-Region Methods: Steihaug, BOBYQA and NEWUOA"
author: "Matthew Fidler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{More Trust-Region Methods: Steihaug, BOBYQA and NEWUOA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Besides `trust()`, `RcppTrust` ships three more trust-region optimizers.
Each one has the same two layers as `trust()`: an R function for
interactive use, and a thread-safe C core with no R API calls and no
global state. The C core can run as an inner or outer optimizer inside
parallel C++ code such as `nlmixr2est`.

| R function | Ported from | Derivatives needed | Bounds | Thread-safe C entry point |
|---|---|---|---|---|
| `trust()` | CRAN `trust` (Geyer) | gradient + Hessian | no | `trust_solve_c()` |
| `steihaug()` | Rust crate [basin](https://github.com/jolars/basin) | gradient + Hessian, or Hessian-vector products | no | `steihaug_solve_c()` |
| `bobyqa()` | CRAN `minqa` (Powell's Fortran) | none | box | `bobyqa_solve_c()` |
| `newuoa()` | CRAN `minqa` (Powell's Fortran) | none | no | `newuoa_solve_c()` |

## How the methods differ

All four repeat the same loop: build a quadratic model of the
objective, minimize it within a radius around the current point,
compare the actual decrease with the predicted one, and grow or shrink
the radius. They differ in where the model comes from and how the
subproblem is solved.

- **`trust()`** uses the exact gradient and Hessian, and solves the
  subproblem nearly exactly with an eigendecomposition. That is robust,
  but each iteration costs O(n^3).
- **`steihaug()`** uses the same derivative-based model, but solves the
  subproblem approximately by conjugate gradients (CG). CG stops early
  when it reaches the trust-region boundary or finds a direction of
  negative curvature. The Hessian is never factorized. If you supply
  only Hessian-vector products (`hessvec`), the Hessian is never formed
  at all.
- **`bobyqa()` and `newuoa()`** never use derivatives. They fit the
  quadratic model through `npt` previously evaluated points (usually
  `2n + 1`), and take up the model's remaining freedom with the
  smallest change to its Hessian. The step is found by truncated CG, so
  NEWUOA's subproblem solver is close in spirit to Steihaug's; BOBYQA's
  also respects box bounds. They track two radii: `rho`, a lower bound
  that only shrinks, from `rhobeg` to `rhoend`, and controls
  termination; and `delta`, the step radius, which can grow and shrink
  above it.

## Steihaug truncated-CG trust region

`steihaug()` takes separate objective, gradient and Hessian functions,
because basin requests each one separately. It reproduces basin's
evaluation pattern:

- `fn` and `gr` once at the start;
- `hess` once per outer iteration;
- `fn` at each trial point;
- `gr` after each accepted step.

```{r}
library(RcppTrust)

fr <- function(x) 100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
grr <- function(x) c(-400 * x[1] * (x[2] - x[1]^2) - 2 * (1 - x[1]),
                     200 * (x[2] - x[1]^2))
hr <- function(x) matrix(c(1200 * x[1]^2 - 400 * x[2] + 2, -400 * x[1],
                           -400 * x[1], 200), 2, 2)

r <- steihaug(c(-1.2, 1), fr, grr, hr)
r[c("par", "value", "iterations", "message", "counts")]
```

In matrix-free mode you supply `hessvec(x, v)` instead of `hess`. The
iterates are the same, but CG calls `hessvec` once per product instead
of forming the matrix:

```{r}
hv <- function(x, v) drop(hr(x) %*% v)
m <- steihaug(c(-1.2, 1), fr, grr, hessvec = hv)
identical(m$par, r$par)
m$counts
```

`control$trace = TRUE` returns one row per subproblem attempt, with the
radius, predicted and actual reduction, `rho`, whether the step was
accepted, whether it hit the boundary, and how many CG iterations it
took:

```{r}
tr <- steihaug(c(-1.2, 1), fr, grr, hr, control = list(trace = TRUE))$trace
head(tr)
```

### What matches basin

The C++ port reproduces basin's arithmetic in the same order, with
floating-point contraction turned off. On 33 test setups, each run in
both exact and matrix-free mode, it matches a basin reference program
bit for bit: every evaluation point, iteration count, termination
reason and evaluation count.

The differences from basin are deliberate:

- **Defaults:** `maxit` is 100 (basin's executor allows 1000) and the
  absolute gradient tolerance is on at `1e-8` (off in basin). Every
  other tolerance is off unless you set it in `control`, as in basin.
- **Errors:** an error in an R callback is raised again as the same R
  error. In C, the solver returns an error code together with the state
  at the start of the failing iteration, where basin throws the state
  away.
- **Not ported:** basin's wall-clock limits, observers and cancellation
  tokens.

## BOBYQA and NEWUOA

`bobyqa()` and `newuoa()` are drop-in replacements for `minqa::bobyqa()`
and `minqa::newuoa()`. The arguments, `control` settings, defaults,
warnings, printed output and return values are the same:

```{r}
fr <- function(x) 100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
bobyqa(c(1, 2), fr, lower = c(0, 0), upper = c(4, 4))
newuoa(c(-1.2, 1), fr)
```

The Fortran was translated line by line into C++. The port keeps
Powell's variable names, workspace layout, `goto` control flow and
order of operations. Compared with `minqa`, the parameters, objective
values, evaluation counts and error codes are bitwise identical:

```{r, eval = requireNamespace("minqa", quietly = TRUE)}
extRosen <- function(x) {
  n <- length(x)
  sum(100 * (x[-1] - x[-n]^2)^2 + (1 - x[-n])^2)
}
ours <- bobyqa(rep(0.5, 6), extRosen, lower = -2, upper = 2,
               control = list(npt = 13))
theirs <- minqa::bobyqa(rep(0.5, 6), extRosen, lower = -2, upper = 2,
                        control = list(npt = 13))
identical(unclass(ours), unclass(theirs))
```

**One intentional difference.** In `minqa` 1.2.8, BOBYQA's `RESCUE`
routine calls the objective as `CALFUN(N,X,IPRINT)`, but `X` is not
defined inside `RESCUE`. As a result, `minqa` evaluates the objective
at whatever is in uninitialized memory, and the same run can give
different results from one call to the next. The port evaluates the
point Powell's original code intended, `W(1..N)`. `RESCUE` is only
entered on badly conditioned problems, so ordinary runs never hit it.
The runs that do enter `RESCUE` were checked against a copy of `minqa`
with only that line fixed, and matched it exactly.

UOBYQA from `minqa` is not included. It needs `(n+1)(n+2)/2` points,
which grows as O(n^2), and NEWUOA covers the same unconstrained,
derivative-free case with far fewer points.

## The thread-safe C interfaces

Each solver has a plain C header, installed with the package:

- `minqa_types.h`: `bobyqa_solve_c()`, `newuoa_solve_c()`,
  `minqa_result_free()`, `minqa_options_default()`;
- `steihaug_types.h`: `steihaug_solve_c()`, `steihaug_result_free()`,
  `steihaug_options_default()`.

The conventions are the same as for `trust_solve_c()`:

- Options live in a plain struct.
- The objective receives an opaque `userdata` pointer.
- Results are written to a caller-supplied struct that owns its buffers
  until you call the matching `*_result_free()`.
- There are no locks and no shared mutable state, so any number of
  threads can each run their own solve.

For the minqa solvers, printing (`iprint`) goes through an optional
callback instead of `Rprintf()`, and is silent by default.

A consuming package reaches these cores through the same header-only
pointer table described in `vignette("RcppTrust")`. After
`iniRcppTrustPtrs()` runs, `bobyqa_solve_c_ptr`, `newuoa_solve_c_ptr`,
`minqa_result_free_ptr`, `steihaug_solve_c_ptr` and
`steihaug_result_free_ptr` are available alongside `trust_solve_c_ptr`.
New slots are only ever appended to the table. When the installed
`RcppTrust` is older than the header a consumer was built against, the
missing pointers are left `NULL`, so check them before use.

The example below runs NEWUOA and matrix-free Steihaug on many starting
points inside an OpenMP loop, through the resolved pointers:

```{r}
cpp_code <- '
// [[Rcpp::depends(RcppTrust)]]
#include <Rcpp.h>
#include <vector>

extern "C" {
#define iniRcppTrustPtrs _vignette2_iniRcppTrustPtrs
#include <RcppTrust.h>
iniRcppTrust
}

// Derivative-free objective for NEWUOA.
extern "C" int rosen_f(int n, const double *x, double *f, void *) {
  double t = x[1] - x[0] * x[0];
  *f = 100.0 * t * t + (1.0 - x[0]) * (1.0 - x[0]);
  return 0;
}

// Value/gradient on request, for Steihaug. The Hessian is never requested
// in matrix-free mode.
extern "C" int rosen_vg(int n, const double *x, double *value,
                        double *gradient, double *hessian, void *) {
  double t = x[1] - x[0] * x[0];
  if (value) *value = 100.0 * t * t + (1.0 - x[0]) * (1.0 - x[0]);
  if (gradient) {
    gradient[0] = -400.0 * x[0] * t - 2.0 * (1.0 - x[0]);
    gradient[1] = 200.0 * t;
  }
  return 0;
}

extern "C" int rosen_hv(int n, const double *x, const double *v, double *hv,
                        void *) {
  double h11 = 1200.0 * x[0] * x[0] - 400.0 * x[1] + 2.0, h12 = -400.0 * x[0];
  hv[0] = h11 * v[0] + h12 * v[1];
  hv[1] = h12 * v[0] + 200.0 * v[1];
  return 0;
}

// [[Rcpp::export]]
Rcpp::NumericMatrix fit_many(SEXP ptrTable, int nStarts) {
  _vignette2_iniRcppTrustPtrs(ptrTable);  // normally done once in .onLoad()
  Rcpp::NumericMatrix out(nStarts, 4);
  std::vector<double> res(4 * nStarts);
  // Build with OpenMP (e.g. -fopenmp in the package Makevars) to run
  // this loop in parallel; without it the pragma is ignored.
  #pragma omp parallel for num_threads(2)
  for (int i = 0; i < nStarts; i++) {
    double start[2] = {-1.2 + 0.05 * i, 1.0};

    minqa_options_t mo = minqa_options_default(2, start);
    minqa_result_t mr;
    newuoa_solve_c_ptr(2, start, rosen_f, nullptr, &mo, &mr);

    steihaug_options_t so = steihaug_options_default();
    steihaug_result_t sr;
    steihaug_solve_c_ptr(2, start, rosen_vg, rosen_hv, nullptr, &so, &sr);

    res[4 * i] = mr.par[0];
    res[4 * i + 1] = mr.feval;
    res[4 * i + 2] = sr.argument[0];
    res[4 * i + 3] = sr.iterations;
    minqa_result_free_ptr(&mr);
    steihaug_result_free_ptr(&sr);
  }
  for (int i = 0; i < nStarts; i++)
    for (int j = 0; j < 4; j++) out(i, j) = res[4 * i + j];
  Rcpp::colnames(out) = Rcpp::CharacterVector::create(
      "newuoa_x1", "newuoa_feval", "steihaug_x1", "steihaug_iter");
  return out;
}
'
Rcpp::sourceCpp(code = cpp_code)
head(fit_many(RcppTrust:::.RcppTrustPtr(), 16L))
```

Every call inside the loop gets its own options, result and workspace,
so the threads need no coordination. The package's test suite runs each
solver on 64 starting points both sequentially and across OpenMP
threads, and checks that the two sets of results are bitwise identical.
